% Journal of Emerging Investigators (JEI) Submission
% Template for high school research paper

\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{lineno}

% JEI requires line numbers
\linenumbers

\title{Modeling the Celiac Gut-Brain Axis with Heterogeneous Graph Neural Networks}

\author{
    [Author Name]\\
    [School Name]\\
    [City, State]\\
    \texttt{[email]}
}

\date{}

\begin{document}

\maketitle

% =============================================================================
\begin{abstract}
Celiac disease (CeD) affects approximately 1\% of the global population and is increasingly associated with neurological manifestations including cerebellar ataxia, peripheral neuropathy, and cognitive dysfunction. However, the biological mechanisms connecting intestinal inflammation to brain pathology remain poorly understood. We present a computational approach using heterogeneous graph neural networks (GNNs) to model the celiac gut-brain axis. Our knowledge graph integrates celiac-associated gut microbiome alterations, microbial metabolites (short-chain fatty acids, tryptophan derivatives), immune signaling genes, and neurological phenotypes from the Human Phenotype Ontology. Using link prediction, our model achieves an AUROC of 0.958 in predicting gene-phenotype associations, identifying candidate biological pathways that may mediate gut-brain communication in celiac disease. This work demonstrates how graph-based machine learning can generate testable hypotheses for complex multi-system diseases.
\end{abstract}

\textbf{Keywords:} celiac disease, gut-brain axis, graph neural networks, knowledge graph, microbiome, neurological manifestations

% =============================================================================
\section{Introduction}

Celiac disease (CeD) is an autoimmune disorder triggered by dietary gluten in genetically susceptible individuals, primarily those carrying HLA-DQ2 or HLA-DQ8 haplotypes \citep{lebwohl2018celiac}. While traditionally characterized as an intestinal disease causing villous atrophy and malabsorption, CeD is increasingly recognized as a systemic condition with diverse extra-intestinal manifestations \citep{caio2019celiac}.

Neurological complications occur in 10-22\% of celiac patients and can include:
\begin{itemize}
    \item Gluten ataxia (cerebellar dysfunction)
    \item Peripheral neuropathy
    \item Cognitive impairment ("brain fog")
    \item Headaches and migraines
    \item Depression and anxiety
\end{itemize}

The mechanisms connecting intestinal inflammation to neurological symptoms remain unclear. Proposed pathways include:
\begin{enumerate}
    \item \textbf{Immune-mediated:} Cross-reactive antibodies (anti-TG6) targeting brain tissue
    \item \textbf{Microbiome-mediated:} Altered gut bacteria producing neuroactive metabolites
    \item \textbf{Nutrient deficiency:} Malabsorption of B vitamins and other nutrients
\end{enumerate}

Understanding these connections requires integrating diverse biological data types. Knowledge graphs provide a natural framework for representing heterogeneous biomedical relationships, and graph neural networks (GNNs) can learn from these complex structures to predict novel associations \citep{zhang2022graph}.

In this study, we construct a knowledge graph representing the celiac gut-brain axis and apply heterogeneous GNNs to predict gene-phenotype associations. Our approach identifies candidate pathways connecting intestinal dysbiosis to neurological manifestations.

% =============================================================================
\section{Methods}

\subsection{Knowledge Graph Construction}

We constructed a heterogeneous knowledge graph with four node types and multiple edge types (Figure~\ref{fig:kg_schema}):

\textbf{Node Types:}
\begin{itemize}
    \item \textbf{Microbe} (n=14): Celiac-associated gut bacteria from meta-analyses \citep{arcila2022microbiome}
    \item \textbf{Metabolite} (n=12): Neuroactive compounds including short-chain fatty acids, tryptophan metabolites
    \item \textbf{Gene} (n=113): Immune, neurotransmitter, and metabolic genes
    \item \textbf{Phenotype} (n=86): Human Phenotype Ontology terms for neurological symptoms
\end{itemize}

\textbf{Edge Types:}
\begin{itemize}
    \item Microbe $\rightarrow$ Metabolite: produces, metabolizes, modulates
    \item Metabolite $\rightarrow$ Gene: activates, substrate\_of, transported\_by
    \item Gene $\rightarrow$ Phenotype: associated\_with (from Monarch Initiative)
\end{itemize}

Gene-phenotype associations were obtained from the Monarch Initiative knowledge base, which aggregates phenotype data from OMIM, ClinVar, and other sources.

\subsection{Graph Neural Network Architecture}

We implemented a heterogeneous GNN using PyTorch Geometric \citep{fey2019pyg}. The architecture consists of:

\begin{enumerate}
    \item \textbf{Node Embeddings:} Learnable embedding vectors (dimension=32) for each node
    \item \textbf{Message Passing:} Two layers of GraphSAGE convolutions \citep{hamilton2017graphsage} with separate weights per edge type
    \item \textbf{Link Prediction:} Bilinear decoder scoring potential gene-phenotype edges
\end{enumerate}

The model was trained to predict held-out gene-phenotype links using binary cross-entropy loss:
\begin{equation}
    \mathcal{L} = -\sum_{(i,j) \in E^+} \log \sigma(s_{ij}) - \sum_{(i,j) \in E^-} \log(1 - \sigma(s_{ij}))
\end{equation}
where $s_{ij}$ is the predicted score for edge $(i,j)$, $E^+$ are positive edges, and $E^-$ are negative samples.

\subsection{Experimental Setup}

We split gene-phenotype edges into train (70\%), validation (15\%), and test (15\%) sets using random link splitting. Negative edges were sampled at a 1:1 ratio. Models were trained for up to 100 epochs with early stopping (patience=20) based on validation AUROC.

% =============================================================================
\section{Results}

\subsection{Knowledge Graph Statistics}

Our final knowledge graph contains 225 nodes and 229 edges (Table~\ref{tab:kg_stats}).

\begin{table}[h]
\centering
\caption{Knowledge Graph Statistics}
\label{tab:kg_stats}
\begin{tabular}{lc}
\toprule
\textbf{Component} & \textbf{Count} \\
\midrule
Microbe nodes & 14 \\
Metabolite nodes & 12 \\
Gene nodes & 113 \\
Phenotype nodes & 86 \\
\midrule
Total nodes & 225 \\
Total edges & 229 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Link Prediction Performance}

The heterogeneous GNN achieved strong performance on the gene-phenotype link prediction task (Table~\ref{tab:results}).

\begin{table}[h]
\centering
\caption{Link Prediction Results}
\label{tab:results}
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{Validation} & \textbf{Test} \\
\midrule
AUROC & 0.987 & 0.958 \\
AUPRC & 0.985 & 0.906 \\
\bottomrule
\end{tabular}
\end{table}

Training converged quickly, with early stopping triggered at epoch 33 (best model at epoch 13). Figure~\ref{fig:training} shows the training curves.

\subsection{Pathway Analysis}

% TODO: Add analysis of top predicted gene-phenotype pairs
% and biological interpretation

% =============================================================================
\section{Discussion}

Our results demonstrate that heterogeneous GNNs can effectively model the complex relationships in the celiac gut-brain axis. The high prediction accuracy (AUROC=0.958) suggests the model captures meaningful biological signal.

\textbf{Limitations:}
\begin{itemize}
    \item Small graph size limits generalization
    \item Curated edges may introduce bias
    \item Model predicts associations, not causation
\end{itemize}

\textbf{Future Directions:}
\begin{itemize}
    \item Incorporate gene expression data from celiac biopsies
    \item Expand to predict microbe-phenotype links
    \item Validate top predictions experimentally
\end{itemize}

% =============================================================================
\section{Conclusion}

We developed a graph-based machine learning approach to model the celiac gut-brain axis. By integrating microbiome, metabolite, gene, and phenotype data into a knowledge graph, our heterogeneous GNN identifies candidate pathways connecting intestinal dysbiosis to neurological manifestations. This computational framework generates testable hypotheses for understanding gut-brain communication in celiac disease.

% =============================================================================
\section*{Acknowledgments}
% TODO: Add acknowledgments

\section*{Data Availability}
Code and data are available at: [GitHub repository URL]

% =============================================================================
\bibliographystyle{plainnat}
\bibliography{references}

% Placeholder references (replace with .bib file)
\begin{thebibliography}{9}

\bibitem{lebwohl2018celiac}
Lebwohl, B., Sanders, D. S., \& Green, P. H. (2018).
Coeliac disease.
\textit{The Lancet}, 391(10115), 70-81.

\bibitem{caio2019celiac}
Caio, G., Volta, U., Sapone, A., et al. (2019).
Celiac disease: a comprehensive current review.
\textit{BMC Medicine}, 17(1), 142.

\bibitem{arcila2022microbiome}
Arcila-Galvis, J. E., et al. (2022).
Systematic review of the gut microbiome in celiac disease.
\textit{Gut Microbes}, 14(1), 2087291.

\bibitem{zhang2022graph}
Zhang, Z., et al. (2022).
Graph neural networks and their applications in drug discovery.
\textit{Expert Opinion on Drug Discovery}, 17(1), 1-13.

\bibitem{fey2019pyg}
Fey, M., \& Lenssen, J. E. (2019).
Fast graph representation learning with PyTorch Geometric.
\textit{ICLR Workshop on Representation Learning on Graphs}.

\bibitem{hamilton2017graphsage}
Hamilton, W., Ying, Z., \& Leskovec, J. (2017).
Inductive representation learning on large graphs.
\textit{NeurIPS}, 30.

\end{thebibliography}

\end{document}
