\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
%  OncoEvidence -- final project deliverable (completed-project report)
%  Compile with:  tectonic docs/deliverable.tex
%  Numbers are pulled from results/*.json in this repository.
% ---------------------------------------------------------------------------

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,shapes.geometric}
\usepackage{booktabs}
\usepackage{array}
\usepackage{colortbl}
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{float}
\graphicspath{{../figures/}{figures/}}

% ----- Palette -------------------------------------------------------------
\definecolor{sriblue}{HTML}{1F4E79}
\definecolor{sriaccent}{HTML}{E8684A}
\definecolor{srigreen}{HTML}{2E8B57}
\definecolor{srigray}{HTML}{4B5563}
\definecolor{srilight}{HTML}{EAF1F8}
\definecolor{sriamber}{HTML}{F6BD16}
\definecolor{srirow}{HTML}{F2F6FB}
\definecolor{srired}{HTML}{C0392B}

\usepackage[colorlinks=true,linkcolor=sriblue,citecolor=sriblue,urlcolor=sriaccent]{hyperref}

\usepackage{titlesec}
\titleformat{\section}{\large\bfseries\color{sriblue}}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{sriblue!85!black}}{\thesubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{1.4ex}{0.8ex}

\newtcolorbox{problembox}{colback=srilight,colframe=sriblue,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{winbox}{colback=sriaccent!8,colframe=sriaccent,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{nullbox}{colback=srigray!8,colframe=srigray,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}

\renewcommand{\arraystretch}{1.25}
\newcommand{\NN}{\mathcal{N}}

\begin{document}

\begin{center}
  {\LARGE\bfseries\color{sriblue} OncoEvidence: Mechanism-Guided Evidence\\ Triage for Cancer Drug Repurposing}\\[4pt]
  {\large\bfseries\color{sriblue!80!black} A completed-project report: when a knowledge graph earns its place,\\ and how to show its mechanism reasoning is predictive and faithful}\\[10pt]
  {\large Akshatha Arunkumar}\\[2pt]
  {\small Final Project Deliverable \quad\textbar\quad \today}
\end{center}

\vspace{2pt}
\begin{problembox}
\textbf{Summary.} OncoEvidence ranks oncology drug-repurposing candidates on the
PrimeKG knowledge graph, extracts the multi-hop mechanism path that would explain
each one, and uses a literature-grounded language model to grade whether that
mechanism is supported. The project is built around an honest question: once a
strong tabular ranker is in place, what does the graph actually add? The answer is
mechanism, and this report goes further than earlier drafts by showing that the
graph's mechanism signal is (i) \emph{predictive} over time and (ii) \emph{faithful}
to curated biology, with one orthogonal validation reported as an honest negative.
\end{problembox}

\begin{winbox}
\textbf{Six findings.}
\begin{enumerate}[leftmargin=1.4em,itemsep=1pt,topsep=2pt]
\item On plain link ranking, a tuned XGBoost matches or beats the graph network in
every regime, so ranking alone does not justify the graph.
\item Only the graph produces a traceable multi-hop mechanism path.
\item Mechanism structure separates true oncology indications from random pairs at
AUROC 0.879, holds up under hard negatives, and agrees with curated DrugMechDB at 0.802.
\item Blinded mechanism recovery: with the direct drug-target edge hidden, only the
jointly trained GNN names the curated bridge gene (Recall@10 0.25), while trivial and
link-only baselines collapse to about zero.
\item \textbf{New: prospective validation.} On a time-split test, the GNN ranks
\emph{future} oncology indications at AUROC 0.930 and beats a structure-blind control
by +0.047 AUROC.
\item \textbf{New: faithfulness.} Deleting a pair's curated mechanism edge drops the
bridge gene's score far more than deleting a random edge (contrast 0.124,
$p \approx 3.6\times 10^{-34}$, faithful in 83\% of cases).
\end{enumerate}
\end{winbox}

% ===========================================================================
\section{Background and significance}

Developing a new drug takes many years and large cost, while thousands of approved
drugs already have known safety profiles. If an approved drug acts on a protein or
pathway that also drives a cancer, repurposing it can be a fast route to a new
treatment. Biomedical knowledge graphs such as PrimeKG encode millions of
relationships among drugs, proteins, genes, pathways, and diseases, so the
information needed to reason about such mechanisms exists as paths through the graph.
The open problem is \emph{trust}: a bare drug-disease score hides its reasoning and
cannot separate a real mechanism from a coincidence.

\textbf{Novelty, stated carefully.} Knowledge-graph repurposing with explanations is
not new: TxGNN is a graph foundation model with multi-hop rationales, and KGML-xDTD
predicts treatment links with KG-based mechanism paths. This project does not claim
``a knowledge graph plus a language model for repurposing'' as novel. The defensible
contribution is methodological: quantify \emph{when} graph structure adds value
beyond a strong tabular ranker, show that its value is mechanism extraction rather
than better ranking, and then audit those mechanisms three ways that most demos skip:
against curated mechanism databases, against a prospective time-split, and against a
counterfactual faithfulness test.

% ===========================================================================
\section{Datasets and system}

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.22\linewidth} >{\raggedright\arraybackslash}p{0.50\linewidth} >{\raggedright\arraybackslash}p{0.20\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Resource} & \color{white}\textbf{Content} & \color{white}\textbf{Role}\\
\midrule
\rowcolor{srirow} PrimeKG & $\sim$129k nodes, $\sim$8.1M directed edges; 7{,}957 drugs, 17{,}080 diseases ($\sim$3{,}170 oncology) & Ranking + mechanism paths\\
Europe PMC & abstracts and open-access full text via REST & Evidence + time axis\\
\rowcolor{srirow} DrugMechDB & curated drug mechanism-of-action paths (UniProt-keyed) & Mechanism ground truth\\
ClinicalTrials.gov & registered interventional studies (API v2) & Orthogonal check\\
\bottomrule
\end{tabular}
\caption{Public resources. DrugMechDB UniProt accessions are mapped to HGNC symbols
via mygene.info; the per-pair time axis in Finding 5 is the earliest Europe PMC
co-mention year.}
\label{tab:data}
\end{table}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=5mm and 8mm,
  box/.style={rounded corners,draw=sriblue,fill=srilight,thick,align=center,font=\footnotesize,inner sep=3pt,minimum height=0.85cm},
  acc/.style={rounded corners,draw=sriaccent,fill=sriaccent!10,thick,align=center,font=\footnotesize,inner sep=3pt,minimum height=0.85cm},
  grn/.style={rounded corners,draw=srigreen,fill=srigreen!10,thick,align=center,font=\footnotesize,inner sep=3pt,minimum height=0.85cm},
  >={Stealth[length=2.2mm]}]
  \node[box] (kg) {PrimeKG\\graph};
  \node[acc,right=of kg] (rank) {Candidate\\ranking\\(GNN / XGBoost)};
  \node[box,right=of rank] (mech) {Mechanism\\path};
  \node[grn,right=of mech] (llm) {LLM evidence\\verifier};
  \node[box,right=of llm] (out) {Ranked\\dossier};
  \draw[->,thick] (kg)--(rank); \draw[->,thick] (rank)--(mech);
  \draw[->,thick] (mech)--(llm); \draw[->,thick] (llm)--(out);
  \node[srigray,font=\scriptsize\itshape,below=2mm of mech] {Europe PMC abstracts feed the verifier};
\end{tikzpicture}
\caption{The OncoEvidence pipeline: rank candidates, extract a mechanism path,
verify it against the literature, output an evidence-backed shortlist.}
\label{fig:pipeline}
\end{figure}

% ===========================================================================
\section{Methods}

\textbf{Ranking.} Repurposing is link prediction on drug-indication-disease edges. A
heterogeneous GraphSAGE encoder builds node embeddings by message passing,
\begin{equation}
h_v^{(l)} = \sigma\!\Big( W^{(l)}\big[\,h_v^{(l-1)}\ \big\Vert\
\textstyle\operatorname*{AGG}_{u\in\NN(v)} h_u^{(l-1)}\big]\Big),
\qquad s(d,c) = \mathrm{MLP}\big([z_d \Vert z_c]\big).
\label{eq:gnn}
\end{equation}
The GNN, a DistMult embedding, and a tuned XGBoost all receive the \emph{same}
SentenceTransformer node features, so any gap measures graph structure, not features.

\textbf{Mechanism extraction.} For a candidate we search PrimeKG for short paths
through mechanism relations, scoring each by specificity and penalizing promiscuous
hubs:
\begin{equation}
\mathrm{score}(p) = \beta_{\text{type}} + \frac{1}{\log_2(\deg(\text{bridge})+2)}\cdot
\frac{1}{\log_2(\delta_{\text{drug}}+2)},
\label{eq:mech}
\end{equation}
where $\delta_{\text{drug}}$ is the number of drugs that bind the bridge protein (an
inverse-frequency penalty that suppresses carriers such as albumin) and
$\beta_{\text{type}}$ orders direct-target over interaction over shared-pathway.

\textbf{Evidence verification.} For the top path we retrieve abstracts, extract
sentences mentioning both drug and gene, and ask a language model to grade the
mechanism supported / weak / contradicted / unknown, requiring a verbatim quote and
grading a purely statistical association as weak.

\textbf{Mechanism head.} A joint objective adds a contrastive term that ranks the
curated DrugMechDB bridge gene above degree-matched decoys, using the same encoder
embeddings, $s_{\text{mech}}(d,g,c) = \mathrm{MLP}([z_d \Vert z_g \Vert z_c])$. This
is the head used in Findings 4 and 6.

% ===========================================================================
\section{Results}

\subsection{Finding 1: the link task does not need the graph}
With a tuned XGBoost on shared features, the GNN does not win in any regime
(Table~\ref{tab:bench}). For link ranking with rich text features, a strong tabular
model suffices; a drug/disease name embedding lets it take a semantic-similarity
shortcut without graph reasoning.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.34\linewidth} cccc}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Test AUROC (5 seeds)} & \color{white}\textbf{GNN} & \color{white}\textbf{XGBoost} & \color{white}\textbf{MLP} & \color{white}\textbf{KGE}\\
\midrule
\rowcolor{srirow} Transductive & 0.977 & \textbf{0.988} & 0.973 & 0.907\\
Inductive (cold-disease, oncology) & 0.882 & \textbf{0.963} & 0.930 & 0.452\\
\rowcolor{srirow} Inductive (cold-drug) & 0.956 & \textbf{0.958} & 0.931 & 0.482\\
\bottomrule
\end{tabular}
\caption{Corrected benchmark. The DistMult memorizer (KGE) collapses on unseen nodes,
confirming the task needs content features, not memorized identity.}
\label{tab:bench}
\end{table}

\subsection{Finding 2: the graph carries real mechanism}
A tabular model cannot produce a traceable mechanism; the graph can. On true
oncology indications the extractor recovers textbook direct-target mechanisms, for
example Quizartinib-FLT3 (myeloid leukemia), Tamibarotene-RARA (acute promyelocytic
leukemia), Topotecan-TOP1 (ovarian cancer), and Trifluridine-TYMS (colorectal cancer).

\subsection{Finding 3: mechanism structure separates true from random pairs}
Over 400 true indications versus 400 random pairs, the mechanism signal separates the
two groups at \textbf{AUROC 0.879} (Table~\ref{tab:sep}). The extracted bridge genes
also agree with curated DrugMechDB mechanisms for 211 of 263 covered pairs
(\textbf{agreement 0.802}). The language-model verifier is the conservative reviewer:
of its ``supported'' calls that DrugMechDB covers, 6/7 are curated-confirmed, versus
13/22 for a keyword baseline (small denominators, reported as a pilot signal).

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.40\linewidth} cc}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Metric} & \color{white}\textbf{True} & \color{white}\textbf{Random}\\
\midrule
\rowcolor{srirow} Mean mechanism score & \textbf{1.95} & 0.17\\
Direct-target rate & \textbf{34.0\%} & 0.25\%\\
\rowcolor{srirow} Any mechanistic path & \textbf{81.3\%} & 8.8\%\\
\bottomrule
\end{tabular}
\quad
\raisebox{-2.2\height}{\parbox{0.30\linewidth}{\centering\textbf{Separation AUROC}\\[2pt]
{\Large\color{sriaccent}\textbf{0.879}}}}
\caption{Mechanism paths separate true indications from random pairs. A true
indication is roughly $136\times$ more likely to have a direct drug-target to
cancer-gene link.}
\label{tab:sep}
\end{table}

\subsection{Finding 4: the graph's one genuine ranking-free edge}
We trained the GNN jointly and asked it to name the curated bridge gene for held-out
cold-disease pairs (3 seeds, $n\approx186$--$260$). Unblinded, a trivial
``the drug's own targets'' baseline dominates because the bridge gene is usually the
drug's direct target. But with that direct edge hidden (mechanism-blinded), the
trivial and degree baselines collapse to about zero while the joint GNN still recovers
the gene from indirect structure (Table~\ref{tab:mechrec}). A link-only GNN scored
through the identical head recovers nothing, so the signal is attributable to the
auxiliary objective.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.40\linewidth} cc c}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Bridge-gene recovery} & \color{white}\textbf{Unblinded R@10} & \color{white}\textbf{Blinded R@10} & \color{white}\textbf{Blinded MRR}\\
\midrule
\rowcolor{srirow} Joint GNN (link + mechanism) & 0.31 & \textbf{0.25} & \textbf{0.21}\\
Trivial target-lookup & \textbf{0.80} & 0.00 & 0.00\\
\rowcolor{srirow} Link-only GNN (same head, no aux loss) & 0.00 & 0.00 & 0.00\\
Degree prior & 0.02 & 0.02 & 0.02\\
\bottomrule
\end{tabular}
\caption{Mechanism recovery on held-out pairs. When the direct edge is hidden, only
the joint GNN recovers the gene, a small ($\sim$0.25 R@10) but genuine graph-only
signal. Caveat: this is recovery of \emph{known} mechanism via indirect paths, not de
novo discovery, on a modest sample.}
\label{tab:mechrec}
\end{table}

\subsection{Finding 5 (new): the model is predictive over time}
The findings above are retrospective. To test whether the model could have flagged an
indication \emph{before} it was established, we assign each true oncology pair an
approximate first-evidence year (earliest Europe PMC co-mention), choose cutoff
$T=2005$, keep only pairs known by $T$ inside the message-passing graph, and hold out
the later pairs as a prospective test set (350 sampled pairs; 341 years resolved; 240
past, 101 future; 3 seeds). The GNN ranks the held-out future indications at
\textbf{AUROC 0.930} and \textbf{AUPRC 0.707}, and beats the structure-blind control
on every metric (Table~\ref{tab:temporal}). Unlike the plain link task in Finding 1,
the graph adds a small but consistent edge here, widest on the harder metrics
(AUPRC +0.115, recall@100 +0.116).

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} cccc}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Prospective ($T{=}2005$)} & \color{white}\textbf{AUROC} & \color{white}\textbf{AUPRC} & \color{white}\textbf{R@50} & \color{white}\textbf{R@100}\\
\midrule
\rowcolor{srirow} GNN (graph) & \textbf{0.930} & \textbf{0.707} & \textbf{0.393} & \textbf{0.703}\\
MLP (structure-blind) & 0.882 & 0.592 & 0.337 & 0.587\\
\midrule
Graph gain & +0.047 & +0.115 & +0.056 & +0.116\\
\bottomrule
\end{tabular}
\caption{Prospective time-split (mean over 3 seeds). The model ranks future
indications well above chance, and the graph beats the tabular-style control on every
metric and seed. Caveat: the year proxy is the earliest literature co-mention, not a
regulatory date, and this is a single seeded 350-pair sample, so magnitudes are
indicative.}
\label{tab:temporal}
\end{table}

\subsection{Finding 6 (new): the mechanism reasoning is faithful}
Can we trust the mechanism the graph names? We take the trained joint model and, for
each held-out pair whose bridge gene is a real drug-target edge, recompute the gene's
mechanism score with the full graph, with that pair's curated mechanism edge deleted,
and with a random other target edge of the same drug deleted (316 instances, 2 seeds,
no retraining). Removing the curated edge degrades the bridge gene far more than
removing a random one (Table~\ref{tab:faith}): a paired Wilcoxon test gives
$p \approx 3.6\times 10^{-34}$, the model is faithful in 83\% of instances, and the two
populations separate at AUROC 0.835. This turns ``the graph explains'' from a claim
into a measured property.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.40\linewidth} cc}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Counterfactual edge removal} & \color{white}\textbf{Score drop} & \color{white}\textbf{Rank drop}\\
\midrule
\rowcolor{srirow} Remove curated mechanism edge & \textbf{0.124} & \textbf{$\sim$211}\\
Remove random target edge (same drug) & $-0.001$ & $\sim$2\\
\midrule
Contrast (faithfulness) & \textbf{0.124} & ---\\
\bottomrule
\end{tabular}
\caption{Faithfulness by counterfactual edge removal (pooled, 316 instances). Faithful
in 83.3\% of cases; paired Wilcoxon $p\approx3.6\times10^{-34}$; separation AUROC 0.835.
Measured on the well-posed population (pairs whose bridge gene is an actual
drug-target edge), in logit space.}
\label{tab:faith}
\end{table}

\subsection{An honest negative: orthogonal trial validation}
\begin{nullbox}
We also asked whether top \emph{novel} predictions (not already PrimeKG indications)
appear as registered oncology trials more than matched controls. Top raw-score
predictions were not enriched over random (6.7\% versus 7.4\%, Fisher $p=0.69$). Raw
score does track trial existence above chance (AUROC 0.676 over 810 pairs,
$p\approx1.5\times10^{-5}$), but this is a popularity confound: one broadly-indicated
drug (folic acid) accounts for 78\% of the top set's hits. The specificity-lift
ranking that de-confounds popularity shows no enrichment at all (0.55$\times$ versus
random, $p=0.97$). We report this as inconclusive: the raw scores track human trial
activity mostly through popular drugs, not through the specific novel shortlist. A
usable version needs MeSH/ChEBI name normalization and trial phase/recency, which is
future work.
\end{nullbox}

% ===========================================================================
\section{Discussion}

The contribution is a disciplined account of where a knowledge graph helps. On plain
link ranking it does not beat a tuned tabular model, and we say so. Its value is the
traceable mechanism a tabular model cannot produce, and this report strengthens that
case along the two axes that matter for a real triage tool. Finding 5 shows the
mechanism-aware graph model is \emph{predictive}: trained only on pre-2005 structure,
it ranks post-2005 indications at 0.930 AUROC and beats the structure-blind control,
so the signal is not merely retrospective bookkeeping. Finding 6 shows the mechanism
reasoning is \emph{faithful}: deleting the curated edge, and only that edge, collapses
the prediction, so the explanation reflects the biology the model actually used.

\textbf{Limitations.} The blinded recovery and faithfulness tests are recovery of
known mechanism, not de novo discovery; some samples are modest; the time axis is a
literature proxy; and the orthogonal trial check was a negative. None of the
predictions are medical advice.

% ===========================================================================
\section{Reproducibility}

Open tools (PyTorch Geometric, scikit-learn, XGBoost) and public data (PrimeKG,
Europe PMC, DrugMechDB, ClinicalTrials.gov). Every reported number has a script and a
JSON record in \texttt{results/}: \texttt{oncorepurpose.json} (Finding 1),
\texttt{mechanism\_eval.json} (Finding 3), \texttt{mechanism\_recovery\_eval.json}
(Finding 4), \texttt{temporal\_split\_eval.json} (Finding 5),
\texttt{edge\_faithfulness\_eval.json} (Finding 6), and
\texttt{clinicaltrials\_validation.json} (negative). A six-part Colab notebook track
reproduces the pipeline end to end, with Part 4 fully self-contained. All predictions
are hypothesis-generating research, not medical advice.

% ===========================================================================
\begin{thebibliography}{9}
\small
\bibitem{txgnn} Huang K, Chandak P, et al. \textit{A foundation model for
clinician-centered drug repurposing (TxGNN)}. Nature Medicine, 2024.
\bibitem{primekg} Chandak P, Huang K, Zitnik M. \textit{Building a knowledge graph
to enable precision medicine (PrimeKG)}. Scientific Data, 2023.
\bibitem{kgmlxdtd} \textit{KGML-xDTD: a knowledge graph-based framework for drug
treatment prediction and mechanism description}. 2023.
\bibitem{decagon} Zitnik M, Agrawal M, Leskovec J. \textit{Modeling polypharmacy
side effects with graph convolutional networks (Decagon)}. Bioinformatics, 2018.
\bibitem{drugmechdb} Mayers M, et al. \textit{DrugMechDB: a curated database of
drug mechanism paths}. 2022.
\bibitem{sage} Hamilton W, Ying R, Leskovec J. \textit{Inductive representation
learning on large graphs (GraphSAGE)}. NeurIPS, 2017.
\end{thebibliography}

\end{document}
