\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
%  OncoRepurpose-GNN -- SRI Research Proposal
%  Compile with:  tectonic docs/proposal.tex
% ---------------------------------------------------------------------------

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,shapes.geometric}
\usepackage{booktabs}
\usepackage{array}
\usepackage{colortbl}
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[colorlinks=true,linkcolor=sriblue,citecolor=sriblue,urlcolor=sriaccent]{hyperref}

% ----- Palette -------------------------------------------------------------
\definecolor{sriblue}{HTML}{1F4E79}
\definecolor{sriaccent}{HTML}{E8684A}
\definecolor{srigreen}{HTML}{2E8B57}
\definecolor{srigray}{HTML}{4B5563}
\definecolor{srilight}{HTML}{EAF1F8}
\definecolor{sriamber}{HTML}{F6BD16}
\definecolor{srirow}{HTML}{F2F6FB}
\definecolor{srired}{HTML}{C0392B}
\definecolor{nDrug}{HTML}{E8684A}
\definecolor{nDis}{HTML}{5B8FF9}
\definecolor{nGene}{HTML}{2E8B57}
\definecolor{nPath}{HTML}{F6BD16}
\definecolor{nPhen}{HTML}{9B59B6}

\usepackage{titlesec}
\titleformat{\section}{\large\bfseries\color{sriblue}}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{sriblue!85!black}}{\thesubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{1.4ex}{0.8ex}

\newtcolorbox{problembox}{colback=srilight,colframe=sriblue,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{gapbox}{colback=sriaccent!8,colframe=sriaccent,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{aimbox}[1]{colback=srigreen!7,colframe=srigreen,boxrule=0.8pt,
  arc=2pt,left=6pt,right=6pt,top=4pt,bottom=4pt,title={#1},
  fonttitle=\bfseries\color{white},colbacktitle=srigreen}

\renewcommand{\arraystretch}{1.25}
\newcommand{\NN}{\mathcal{N}}

\begin{document}

\begin{center}
  {\LARGE\bfseries\color{sriblue} Finding New Cancer Uses for Existing Drugs}\\[3pt]
  {\large\bfseries\color{sriblue!80!black} A heterogeneous graph neural network for drug repurposing on a\\ biomedical knowledge graph -- and an honest test of when the graph helps}\\[10pt]
  {\large Akshatha Arunkumar}\\[2pt]
  {\small Student Research Institute (SRI) -- Summer 2026 Cohort \\
  Harvard Undergraduate OpenBio Laboratory}\\[2pt]
  {\small\itshape Research Proposal for Mentor Review \quad\textbar\quad \today}
\end{center}

\vspace{2pt}
\begin{problembox}
\textbf{The problem.} Bringing a new cancer drug to patients takes more than a
decade and often fails. \emph{Drug repurposing} -- finding a new therapeutic use
for an already-approved drug -- is far faster and cheaper, and is especially
valuable for cancers with few options. But the space of (drug, cancer) pairs is
enormous and most candidates are dead ends. The key fact is that whether a drug
might treat a cancer is rarely written in the drug's own properties; it lives in
a \emph{web of relationships} -- the drug's protein targets, the pathways those
proteins sit in, and the molecular processes the cancer disrupts. This project
builds a model that reasons over that web to rank promising repurposing
candidates, and -- critically -- tests honestly whether the graph structure is
truly necessary or whether a simple feature-based model does just as well.
\end{problembox}

% ===========================================================================
\section{Background and Significance}

Approved drugs have known safety profiles, so a credible repurposing hypothesis
can reach patients years faster than a new molecule. The clinical precedent is
strong: thalidomide for multiple myeloma, metformin in oncology trials, and
kinase inhibitors crossing between cancer types. The challenge is
\emph{prioritization} -- which of millions of (drug, cancer) pairs deserve
costly experimental follow-up.

Whether a drug can treat a disease is fundamentally a \emph{relational} property:

\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.62\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Why it matters} & \color{white}\textbf{Consequence}\\
\midrule
\rowcolor{srirow} Drug action is mediated by targets \& pathways & The signal is in multi-hop links (drug $\to$ protein $\to$ pathway $\to$ disease), not a drug's own descriptors.\\
Knowledge is fragmented & Integrating drugs, genes, pathways, phenotypes, and diseases into one graph exposes routes invisible to single-table analysis.\\
\rowcolor{srirow} Most cancers have unexploited options & A ranked, evidence-backed shortlist focuses scarce experimental effort.\\
\bottomrule
\end{tabular}
\caption{Drug repurposing is a graph-reasoning problem: the evidence for a
(drug, cancer) link is distributed across the biomedical network.}
\label{tab:significance}
\end{table}

% ===========================================================================
\section{The Unmet Need}

\begin{gapbox}
\textbf{Gap 1 -- Scale.} The (drug $\times$ cancer) hypothesis space is too large
to screen experimentally; computational prioritization is essential.\\[3pt]
\textbf{Gap 2 -- Single-table models miss multi-hop biology.} A model that sees
only a drug's and a disease's own features cannot traverse the
drug$\to$target$\to$pathway$\to$disease chains that explain therapeutic action.\\[3pt]
\textbf{Gap 3 -- Predictions are not actionable on their own.} A bare ranked list
without mechanistic rationale or literature support is hard for a researcher to
trust or triage.
\end{gapbox}

\noindent We exploit a simple observation: \textbf{a (drug, disease) link is a
property of the two nodes' positions in the biomedical network}. The information
absent from a drug's descriptors is present in the graph \emph{around} it
(Fig.~\ref{fig:mechanism}). A model that propagates over the network can
represent ``this drug targets a protein in a pathway this cancer dysregulates'';
a feature-only model must guess.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.95,every node/.style={transform shape},
  >={Stealth[length=2.2mm]},
  drug/.style={circle,fill=nDrug,text=white,minimum size=15pt,inner sep=1pt,font=\scriptsize\bfseries},
  dis/.style={circle,fill=nDis,text=white,minimum size=15pt,inner sep=1pt,font=\scriptsize\bfseries},
  gene/.style={circle,fill=nGene,text=white,minimum size=15pt,inner sep=1pt,font=\scriptsize\bfseries},
  path/.style={circle,fill=nPath,text=black,minimum size=15pt,inner sep=1pt,font=\scriptsize\bfseries}]
  \node[drug] (D) at (0,0) {Drug};
  \node[gene] (G1) at (2.1,0.7) {Tgt};
  \node[path] (P) at (4.2,0) {Path};
  \node[gene] (G2) at (2.1,-0.7) {Tgt};
  \node[dis] (X) at (6.3,0) {Cancer};
  \draw[->,thick] (D)--(G1) node[midway,above,font=\tiny]{targets};
  \draw[->,thick] (D)--(G2);
  \draw[->,thick] (G1)--(P) node[midway,above,font=\tiny]{in};
  \draw[->,thick] (G2)--(P);
  \draw[->,thick] (P)--(X) node[midway,above,font=\tiny]{dysregulated in};
  \draw[->,thick,densely dashed,sriaccent] (D) to[bend right=32] node[midway,below,font=\tiny\itshape,sriaccent]{predicted: treats?} (X);
\end{tikzpicture}
\caption{A repurposing hypothesis is a multi-hop path. The dashed link
(does this drug treat this cancer?) is supported by a chain through shared
targets and pathways -- structure a feature-only model cannot see.}
\label{fig:mechanism}
\end{figure}

% ===========================================================================
\section{Research Aims}

\begin{aimbox}{Aims}
\textbf{Aim 1.} Build a heterogeneous graph neural network that predicts
\emph{drug $\to$ cancer} therapeutic links on a large biomedical knowledge graph
and ranks novel repurposing candidates.\\
\textbf{Aim 2.} Test \emph{honestly} whether graph structure is necessary --
benchmark against a tuned tabular model on identical features, under
leakage-safe inductive splits, with a topology-removal ablation.\\
\textbf{Aim 3.} Make predictions actionable: attach multi-hop mechanistic
rationales and an automated, literature-grounded evidence report for the top
candidates.
\end{aimbox}

\noindent\textbf{Guiding questions.} (Q1) Can therapeutic links be predicted
from a biomedical knowledge graph? (Q2) Does message passing beat a strong
feature-only model, and \emph{in which regime} (familiar vs.\ unseen diseases/
drugs)? (Q3) Are the top predictions mechanistically and bibliographically
defensible?

% ===========================================================================
\section{Data}

\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.26\linewidth} >{\raggedright\arraybackslash}p{0.42\linewidth} >{\raggedright\arraybackslash}p{0.24\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Resource} & \color{white}\textbf{Content} & \color{white}\textbf{Role}\\
\midrule
\rowcolor{srirow} PrimeKG & $\sim$129k nodes (drugs, diseases, genes/proteins, pathways, phenotypes, \ldots), $\sim$8M edges incl.\ indication / contraindication & Primary knowledge graph\\
Oncology subset & $\sim$3{,}000 neoplasm diseases flagged for focused evaluation & Cancer-specific targets\\
\rowcolor{srirow} Europe PMC / PubMed & open literature search API & Evidence retrieval for the report layer\\
\bottomrule
\end{tabular}
\caption{All inputs are public. PrimeKG is purpose-built for precision-medicine
AI, with rich drug--disease therapeutic edges other graphs lack.}
\label{tab:data}
\end{table}

% ===========================================================================
\section{Proposed Methodology}

We represent PrimeKG as a typed graph and pose repurposing as \textbf{link
prediction} on the \emph{drug $\xrightarrow{\text{indication}}$ disease} relation.
Every node is given the \emph{same} content feature -- a sentence-transformer
embedding of its name/description -- so the comparison against non-graph models
isolates the value of structure, not features.

\subsection{Proposed model: a heterogeneous GNN}
A heterogeneous GraphSAGE encoder produces a node embedding by pooling
type-specific messages over $L$ hops,
\begin{equation}
h_v^{(l)} = \sigma\!\Big( \textstyle\sum_{r\in\mathcal{R}} W_r^{(l)}
\operatorname*{AGG}_{u\in\NN_r(v)} h_u^{(l-1)} \Big),
\qquad
\hat{p}(d\!\to\!x) = \sigma\!\big( \phi(h_d, h_x) \big),
\label{eq:gnn}
\end{equation}
and a decoder $\phi$ scores a (drug $d$, disease $x$) pair from their embeddings.
Because $h_d$ and $h_x$ are built from multi-hop neighbourhoods, the model can
encode the drug$\to$target$\to$pathway$\to$disease evidence of
Fig.~\ref{fig:mechanism}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=6mm and 8mm,
  box/.style={rounded corners,draw=sriblue,fill=srilight,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  red/.style={rounded corners,draw=sriaccent,fill=sriaccent!10,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  gray/.style={rounded corners,draw=srigray,fill=gray!8,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  >={Stealth[length=2.2mm]}]
  \node[box] (g) {PrimeKG\\(typed graph)};
  \node[red,right=of g] (enc) {HeteroGNN\\encoder};
  \node[box,right=of enc] (z) {Drug + disease\\embeddings};
  \node[red,right=of z] (dec) {Link\\decoder};
  \node[box,right=of dec] (s) {P(drug treats\\cancer)};
  \draw[->,thick] (g)--(enc); \draw[->,thick] (enc)--(z);
  \draw[->,thick] (z)--(dec); \draw[->,thick] (dec)--(s);
  \node[gray,below=12mm of enc] (mlp) {Baselines:\\XGBoost / MLP};
  \node[box,below=12mm of z] (xf) {Node\\features};
  \draw[->,thick,densely dashed,srigray] (xf)--(mlp);
  \draw[->,thick,densely dashed,srigray] (mlp.east) -- ++(0.7,0) |- (dec.south west);
  \node[srigray,font=\scriptsize\itshape,below=1pt of mlp] {baselines never see the graph};
\end{tikzpicture}
\caption{Pipeline. The GNN (orange) pools each node's KG neighbourhood; the
non-graph baselines (gray) score a pair from its two feature vectors alone.}
\label{fig:pipeline}
\end{figure}

\subsection{Baselines and rigorous evaluation}
\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.22\linewidth} >{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.40\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Model} & \color{white}\textbf{Uses graph?} & \color{white}\textbf{Purpose}\\
\midrule
\rowcolor{srirow} \textbf{HeteroGNN (proposed)} & yes (multi-hop) & repurposing that uses network biology\\
XGBoost & no & strong tabular baseline on pair features\\
\rowcolor{srirow} MLP & no & structure-blind neural control\\
DistMult (KGE) & memorized node embeddings & isolates memorization vs.\ generalization\\
\bottomrule
\end{tabular}
\caption{Every model receives the same node features; only the GNN sees the
graph. The KGE has no representation for unseen nodes by design.}
\label{tab:models}
\end{table}

\noindent\textbf{Leakage-safe splits.} (i) \emph{Transductive} -- random
held-out links; all nodes seen. (ii) \emph{Inductive cold-disease} -- whole
cancers held out (all their drug links removed from training), the
disease-discovery regime. (iii) \emph{Inductive cold-drug} -- whole drugs held
out. \textbf{Metrics:} AUROC, AUPRC, Hits@$K$, MRR as mean$\pm$std over 5 seeds,
with paired significance tests of GNN vs.\ each baseline.

% ===========================================================================
\section{Rigor: Ablations, Interpretation, and an Evidence Report}

\subsection{Is the graph really doing the work? (falsification)}
We run the identical GNN on the intact graph, a degree-preserving \emph{shuffled}
graph, and an \emph{empty} graph. If the advantage is genuinely topological,
destroying structure must collapse the GNN toward the feature-only baseline. A
\emph{relation ablation} (removing drug--protein, disease--protein, or pathway
edges) then identifies which biological links carry the signal.

\subsection{From predictions to an actionable shortlist}
For the top candidates we (i) extract short multi-hop KG paths as mechanistic
rationale, (ii) retrieve supporting/contradicting literature (Europe PMC), and
(iii) use a large language model to draft a structured \emph{evidence dossier}
and an \emph{LLM-as-judge} plausibility score that triages the list. The output
is a ranked, citation-grounded set of cancer repurposing hypotheses.

% ===========================================================================
\section{Alignment with Strong-Research Criteria}

\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.62\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Criterion} & \color{white}\textbf{Commitment}\\
\midrule
\rowcolor{srirow} Real, sizable public data & PrimeKG: $\sim$129k nodes, $\sim$8M edges.\\
Clear medical motivation & Faster, cheaper cancer therapy via repurposing.\\
\rowcolor{srirow} Rigorous evaluation & XGBoost + MLP + KGE baselines; inductive splits; 5 seeds; significance tests.\\
Interpretability $\rightarrow$ discovery & Multi-hop rationales; literature-grounded evidence dossiers.\\
\rowcolor{srirow} Ablations & Topology removal; relation ablation.\\
Translational impact & A ranked, vetted oncology repurposing shortlist.\\
\bottomrule
\end{tabular}
\caption{The project is structured around the elements of strong research.}
\label{tab:formula}
\end{table}

% ===========================================================================
\section{Expected Outcomes}

A proposal states targets, not measured results.

\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.26\linewidth} >{\raggedright\arraybackslash}p{0.66\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Outcome} & \color{white}\textbf{Target}\\
\midrule
\rowcolor{srirow} Accurate repurposing & High AUROC/AUPRC on held-out therapeutic links.\\
Value of structure & GNN $\geq$ strong baselines; advantage clearest for unseen diseases/drugs, where memorization (KGE) collapses toward chance.\\
\rowcolor{srirow} Causal check & Topology-removal ablation collapses the GNN to the feature baseline.\\
Actionable output & A ranked, mechanism- and literature-backed cancer repurposing shortlist.\\
\rowcolor{srirow} Honesty clause & A strong tabular baseline is expected to be competitive; any narrowing of the gap is reported as-is, not overclaimed.\\
\bottomrule
\end{tabular}
\caption{Success is an accurate, rigorously-evaluated, interpretable repurposing
method and a vetted candidate shortlist.}
\label{tab:success}
\end{table}

% ===========================================================================
\section{Eight-Week Timeline}

\begin{table}[h]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.14\linewidth} >{\raggedright\arraybackslash}p{0.78\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Weeks} & \color{white}\textbf{Milestones}\\
\midrule
\rowcolor{srirow} 1--2 & Build PrimeKG $\to$ PyG pipeline; shared node features; oncology subset.\\
3--4 & Implement HeteroGNN + baselines (XGBoost, MLP, KGE); shared trainer.\\
\rowcolor{srirow} 5 & Transductive + inductive (cold-disease/cold-drug) experiments, 5 seeds, significance tests.\\
6 & Ablations: topology removal, relation ablation.\\
\rowcolor{srirow} 7 & Interpretability + agentic evidence report; candidate shortlist.\\
8 & Figures, tables, manuscript; reproducibility package.\\
\bottomrule
\end{tabular}
\caption{Schedule for the eight-week SRI cohort.}
\label{tab:timeline}
\end{table}

% ===========================================================================
\section{Reproducibility and Ethics}

Implementation uses open tools (PyTorch Geometric for the model; PrimeKG and
Europe PMC for data). All code, configs, seeds, and a data manifest will be
released; large artifacts are regenerated from public sources rather than
committed. No human subjects are involved; all inputs are public and
de-identified. Predictions are research-only and hypothesis-generating and do
not constitute clinical guidance.

% ===========================================================================
\section*{Selected References}
\small
\begin{enumerate}[leftmargin=1.4em,itemsep=0pt]
\item Huang, Chandak, et al. \textit{A foundation model for clinician-centered
drug repurposing (TxGNN)}. Nature Medicine, 2024.
\item Chandak, Huang, Zitnik. \textit{Building a knowledge graph to enable
precision medicine (PrimeKG)}. Scientific Data, 2023.
\item Zitnik, Agrawal, Leskovec. \textit{Modeling polypharmacy side effects with
graph convolutional networks (Decagon)}. Bioinformatics, 2018.
\item Hamilton, Ying, Leskovec. \textit{Inductive Representation Learning on
Large Graphs (GraphSAGE)}. NeurIPS, 2017.
\item Yang, Trajanoski, et al. \textit{Drug repurposing: progress, challenges
and recommendations}. (review).
\end{enumerate}

\end{document}
