\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
%  Precision Tissue Mapping from Spatial Transcriptomics Using Graph Learning
%  SRI / Harvard OpenBio Research Proposal
%  Compile with:  tectonic docs/proposal.tex   (or pdflatex twice)
% ---------------------------------------------------------------------------

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,shapes.geometric}
\usepackage{booktabs}
\usepackage{array}
\usepackage{colortbl}
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{float}
\graphicspath{{../figures/}{figures/}}

% ----- Palette -------------------------------------------------------------
\definecolor{sriblue}{HTML}{1F4E79}
\definecolor{sriaccent}{HTML}{E8684A}
\definecolor{srigreen}{HTML}{2E8B57}
\definecolor{srigray}{HTML}{4B5563}
\definecolor{srilight}{HTML}{EAF1F8}
\definecolor{sriamber}{HTML}{F6BD16}
\definecolor{srirow}{HTML}{F2F6FB}
\definecolor{srired}{HTML}{C0392B}
\definecolor{sripurple}{HTML}{7B4FAE}
\definecolor{domA}{HTML}{E8684A}
\definecolor{domB}{HTML}{5B8FF9}
\definecolor{domC}{HTML}{2E8B57}

\usepackage[colorlinks=true,linkcolor=sriblue,citecolor=sriblue,urlcolor=sriaccent]{hyperref}

\usepackage{titlesec}
\titleformat{\section}{\large\bfseries\color{sriblue}}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{sriblue!85!black}}{\thesubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{1.4ex}{0.8ex}

\newtcolorbox{problembox}{colback=srilight,colframe=sriblue,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{gapbox}{colback=sriaccent!8,colframe=sriaccent,boxrule=0.9pt,
  arc=2pt,left=6pt,right=6pt,top=5pt,bottom=5pt}
\newtcolorbox{aimbox}[1]{colback=srigreen!7,colframe=srigreen,boxrule=0.8pt,
  arc=2pt,left=6pt,right=6pt,top=4pt,bottom=4pt,title={#1},
  fonttitle=\bfseries\color{white},colbacktitle=srigreen}

\renewcommand{\arraystretch}{1.25}
\newcommand{\NN}{\mathcal{N}}

\begin{document}

\begin{center}
  {\LARGE\bfseries\color{sriblue} Precision Tissue Mapping from Spatial\\ Transcriptomics Using Graph Learning}\\[4pt]
  {\large\bfseries\color{sriblue!80!black} A neighborhood-aware framework for spatially\\ resolved molecular pathology}\\[10pt]
  {\large Akshatha Arunkumar}\\[2pt]
  {\small Student Research Institute (SRI), Summer 2026 Cohort \\
  Harvard Undergraduate OpenBio Laboratory}\\[2pt]
  {\small\itshape Research Proposal for Mentor Review \quad\textbar\quad \today}
\end{center}

\vspace{2pt}
\begin{problembox}
\textbf{The problem.} Spatial transcriptomics measures gene expression while
keeping track of where each measurement sits in the tissue (a spot or a single
cell, depending on the platform). This lets us ask not only which genes change in
disease, but where in the tissue the change occurs, a central question in
precision oncology and neuropathology \cite{cilento2024}. Before a spatial slide
can be interpreted, however, every measured spot must first be assigned to its
tissue domain (a cortical layer, a tumor region, an immune niche). Today that
assignment relies on slow, subjective manual annotation, or on expression-only
clustering that ignores location and produces noisy, spatially incoherent maps.
This project builds an automated, label-efficient model for the mapping step and
tests whether it has to be spatially aware to work.
\end{problembox}

\begin{aimbox}{Proposed study}
This project will test whether tissue-neighborhood information improves
spatial-domain mapping beyond expression-only machine learning. I will use two
public datasets: the LIBD DLPFC benchmark (12 Visium sections, 47{,}329
expert-annotated spots, 7 cortical domains) and single-cell osmFISH mouse cortex
(33 genes). The main comparison will be between a spatial graph neural network
and two non-graph baselines, XGBoost and an MLP, all trained on the same
expression features. The key test is simple: if the graph is really helping, the
GNN should perform better when it has the true tissue-neighborhood graph than
when the graph is shuffled or removed.
\end{aimbox}

\vspace{4pt}
\noindent\textit{Why this project.} I chose tissue mapping because it sits where
neuroscience, cancer biology, and machine learning meet, and because the question
is concrete: can a model tell which part of a tissue a measurement came from?
My main interest is whether local context changes the answer. A spot can be noisy
on its own, but its neighbors may make its tissue domain much clearer. Testing
that idea carefully is what the project is about.

% ===========================================================================
\section{Background and Significance}

\textbf{Why spatial context is clinically important.} Tissue function depends on
how cells are arranged in space. In the cerebral cortex, six layers (L1--L6)
above the white matter each carry distinct cell types and gene programs. In a
tumor, malignant cells, stroma, blood vessels, and infiltrating immune cells form
spatially structured niches. Disease changes are often spatially localized: tau
pathology in Alzheimer's follows a laminar gradient, schizophrenia shows
layer-specific cortical deficits, and a tumor's response to immunotherapy depends
on where immune cells sit relative to the malignant front. Standard bulk RNA
sequencing averages this away. Spatial transcriptomics (10x Visium, MERFISH,
Xenium, CosMx, Stereo-seq) preserves it, and a 2024 review in the Journal of
Cancer Research and Clinical Oncology argues that this spatial information can
advance precision oncology by capturing tumor heterogeneity, metastasis,
prognosis, and immunotherapy response that bulk profiling misses
\cite{cilento2024}.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.62\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Why it matters} & \color{white}\textbf{Consequence}\\
\midrule
\rowcolor{srirow} Tissue function is spatially organized & Disease changes are region/layer-specific, not diffuse.\\
Spatial omics preserves location & We can finally ask \emph{where} pathology starts.\\
\rowcolor{srirow} But spots must be assigned to a domain first & Domain mapping is an important early step for most downstream spatial analyses.\\
\bottomrule
\end{tabular}
\caption{Accurate spatial-domain mapping is the step that turns a
spatial-transcriptomics slide into interpretable, region-resolved biology, in
both brain disease and cancer.}
\label{tab:significance}
\end{table}

% ===========================================================================
\section{The Computational Gap}

Domain mapping is hard for two complementary reasons, and current practice
addresses neither well.

\begin{gapbox}
\textbf{Gap 1. Manual annotation does not scale.} Expert annotation of a single
section is slow and subjective. A disease cohort spans dozens of sections across
many donors, where hand-labelling becomes the bottleneck.\\[3pt]
\textbf{Gap 2. Per-spot expression is weak and structure-blind.} A single spot's
expression is sparse and dropout-corrupted, so models that classify each spot in
isolation (XGBoost, MLP) reach an accuracy ceiling and yield spatially fragmented
domains.
\end{gapbox}

\noindent The observation behind our approach is straightforward: a spot's domain
is largely a property of its spatial neighbourhood. Adjacent spots usually share
a domain, so the signal missing from one noisy spot is present in the tissue
around it (Fig.~\ref{fig:mechanism}). A model that can use the neighbourhood
should therefore do better than one that cannot, and the size of that gap is a
direct measure of how much the spatial structure is worth.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.6,every node/.style={transform shape}]
  \foreach \i in {0,...,5}{
    \foreach \j in {0,...,5}{
      \pgfmathtruncatemacro{\sum}{\i+\j}
      \ifnum\sum<5
        \definecolor{cc}{HTML}{E8684A}
      \else
        \ifnum\sum<8 \definecolor{cc}{HTML}{5B8FF9}\else\definecolor{cc}{HTML}{2E8B57}\fi
      \fi
      \node[circle,fill=cc,minimum size=10pt,inner sep=0pt] (n\i\j) at (\i,\j) {};
    }
  }
  \node[circle,draw=black,line width=1.1pt,minimum size=12pt,inner sep=0pt] at (2,3) {};
  \foreach \a/\b in {1/3,3/3,2/2,2/4}{\draw[black,line width=1pt] (2,3)--(\a,\b);}
  \node[anchor=west,font=\small] at (6.4,4.6) {\textbf{tissue domains} (colours)};
  \node[anchor=west,font=\small,domA] at (6.4,3.8) {$\bullet$ domain A};
  \node[anchor=west,font=\small,domB] at (6.4,3.1) {$\bullet$ domain B};
  \node[anchor=west,font=\small,domC] at (6.4,2.4) {$\bullet$ domain C};
  \node[anchor=west,font=\small] at (6.4,1.4) {a spot + its};
  \node[anchor=west,font=\small] at (6.4,0.8) {spatial neighbours};
\end{tikzpicture}
\caption{Domains are spatially contiguous, so a spot's label tends to agree with
its neighbours'. A single noisy spot may be ambiguous, but its neighbourhood
reveals its domain, which is the signal a spatial model can use and a per-spot
model cannot.}
\label{fig:mechanism}
\end{figure}

% ===========================================================================
\section{Central Hypothesis and Specific Aims}

\begin{problembox}
\textbf{Central hypothesis.} A spot's tissue domain is encoded not only in its
own gene expression but in its spatial neighbourhood. A graph neural network that
performs tissue-neighbour message passing should therefore map domains more
accurately than XGBoost and MLP models trained on the same expression features,
and removing the spatial graph should erase that advantage.
\end{problembox}

\begin{aimbox}{Specific Aims}
\textbf{Aim 1. Build and benchmark.} Develop a spatial GNN for tissue-domain
mapping and compare it directly against tuned XGBoost and an MLP on the same
features, under leakage-safe splits.\\
\textbf{Aim 2. Test whether the advantage is spatial.} Run a graph-removal test
(intact vs.\ shuffled vs.\ empty graph) on two tissues and technologies, and check
whether the GNN loses its advantage when the structure is destroyed.\\
\textbf{Aim 3. Ground it biologically and extend to cancer.} Recover known domain
marker genes, flag candidate annotation errors, and pilot the same pipeline on
tumor spatial transcriptomics to map microenvironment niches.
\end{aimbox}

% ===========================================================================
\section{Relationship to Prior Work}

GNN-based spatial-domain identification is an established field. SpaGCN
\cite{hu2021}, STAGATE \cite{dong2022}, and GraphST \cite{long2023} all apply
graph neural networks to spatial-omics data and are standard baselines on the
DLPFC benchmark.
This proposal does not claim a new architecture. Those methods are predominantly
unsupervised (graph autoencoders followed by clustering, scored by ARI against
manual labels) and are compared mainly against other clustering methods. My
goal is to ask a complementary and, in this field, under-asked question: how
much does spatial structure actually add over strong non-graph models given the
same features? I will answer it with (i) a supervised, leakage-safe cross-section
protocol (train on whole sections, predict held-out ones, which is the
disease-cohort regime); (ii) tuned XGBoost and MLP baselines on the same features,
which spatial-domain papers rarely report; (iii) a graph-removal test that
separates the spatial signal from model capacity; and (iv) replication across two
technologies (Visium and osmFISH). The deliverable will be a careful estimate of
the value of tissue structure. It is planned as a rigorous benchmark and analysis,
not a state-of-the-art or novel-method claim.

% ===========================================================================
\section{Datasets}

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.24\linewidth} >{\raggedright\arraybackslash}p{0.44\linewidth} >{\raggedright\arraybackslash}p{0.24\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Dataset} & \color{white}\textbf{Content} & \color{white}\textbf{Role}\\
\midrule
\rowcolor{srirow} LIBD DLPFC (10x Visium) \cite{maynard2021,pardo2022} & 12 sections, 47{,}329 spots, expert manual layer labels (L1--L6 + WM) & Primary ground-truth benchmark (cross-section)\\
osmFISH mouse cortex \cite{codeluppi2018} & single-cell resolution, 33 genes, $\sim$4{,}839 cells, region labels & Second technology / generalization\\
\rowcolor{srirow} Controlled synthetic tissue & domains defined purely by neighbourhood; per-spot signal weak by construction & Mechanism isolation / sanity check\\
Tumor ST (Visium/Xenium, stretch) & public solid-tumor sections with niche annotations & Clinical extension (Aim 3)\\
\bottomrule
\end{tabular}
\caption{All data are public and de-identified. LIBD DLPFC is the field-standard
ground truth for spatial-domain methods. osmFISH tests a different assay with far
fewer genes, where per-cell features are weakest.}
\label{tab:data}
\end{table}

% ===========================================================================
\section{Methodology}

\textbf{From tissue to spatial graph.} Each section becomes a graph: nodes are
spots or cells, node features are preprocessed expression (normalize, log1p,
highly-variable genes, then PCA), and each node is connected to its $k$ nearest
spatial neighbours. The 2D coordinates are used only to build the graph, never as
model features, so any GNN benefit is attributable to message passing over tissue
structure rather than to memorizing absolute position.

\subsection{Model: a spatial graph neural network}
The GNN will classify each spot by combining two sources of information: the
spot's own gene-expression profile and a summary of nearby spots in the tissue.
After several rounds of this neighborhood aggregation, the model assigns each
spot a tissue-domain label. This is useful because aggregation can denoise a
sparse per-spot signal, which is the limitation described in Gap~2. I will start
with GraphSAGE because it is a standard inductive GNN, then compare it with GAT
(graph attention, which learns per-neighbor weights) and GCN encoders if time
allows. All models will use PyTorch Geometric.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=6mm and 7mm,
  box/.style={rounded corners,draw=sriblue,fill=srilight,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  red/.style={rounded corners,draw=sriaccent,fill=sriaccent!10,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  gray/.style={rounded corners,draw=srigray,fill=gray!8,thick,minimum height=0.9cm,align=center,font=\footnotesize,inner sep=3pt},
  >={Stealth[length=2.2mm]}]
  \node[box] (g) {Spatial\\graph};
  \node[red,right=of g] (enc) {GNN encoder\\(neighbourhood)};
  \node[box,right=of enc] (z) {Spot\\embeddings};
  \node[red,right=of z] (dec) {Softmax\\head};
  \node[box,right=of dec] (s) {Tissue\\domain};
  \draw[->,thick] (g)--(enc); \draw[->,thick] (enc)--(z);
  \draw[->,thick] (z)--(dec); \draw[->,thick] (dec)--(s);
  \node[gray,below=12mm of enc] (mlp) {Baselines:\\MLP / XGBoost};
  \node[box,below=12mm of z] (xf) {Spot\\expression};
  \draw[->,thick,densely dashed,srigray] (xf)--(mlp);
  \draw[->,thick,densely dashed,srigray] (mlp.east) -- ++(0.8,0) |- (dec.south west);
  \node[srigray,font=\scriptsize\itshape,below=1pt of mlp] {baselines classify each spot in isolation};
\end{tikzpicture}
\caption{Pipeline. The GNN (orange) pools each spot's spatial neighbourhood; the
non-spatial baselines (gray) classify each spot alone from the same features. The
only difference between them is access to the spatial graph.}
\label{fig:pipeline}
\end{figure}

\subsection{Baselines and leakage-safe evaluation}
I will benchmark against tuned XGBoost and an MLP on the same expression features, so
the comparison isolates the value of tissue structure. I will use three splits:
(i) \emph{cross-section}, where whole sections are held out for test (layer
layouts differ per section, so absolute position cannot transfer and only
neighbourhood-relative reasoning generalizes; this is the disease-cohort regime
and our headline test); (ii) \emph{within-section}, where contiguous spatial
blocks are held out; and (iii) \emph{stratified}, a standard transductive node
split for single-section data (osmFISH). In the cross-section disjoint-union
graph, train and test sections are separate connected components, so message
passing cannot leak labels across the split. I will report accuracy, macro-F1
(domains are imbalanced), and the adjusted Rand index (ARI, the field-standard
domain metric), as mean $\pm$ std over seeds.

% ===========================================================================
\section{Planned Evaluation}

The main outcome will be whether adding spatial-neighborhood information improves
domain mapping compared with expression-only models. I will report accuracy,
macro-F1, and adjusted Rand index (ARI). Macro-F1 is important because tissue
domains are not equally sized, and ARI is commonly used in spatial-domain papers.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.26\linewidth} >{\raggedright\arraybackslash}p{0.32\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Question} & \color{white}\textbf{Comparison} & \color{white}\textbf{Evidence I will look for}\\
\midrule
\rowcolor{srirow} Does spatial context help? & GNN vs.\ XGBoost / MLP & Higher macro-F1 and ARI on held-out sections or cells.\\
Does the graph itself matter? & True graph vs.\ shuffled/empty graph & Performance drops when tissue neighborhoods are destroyed.\\
\rowcolor{srirow} Does the method generalize? & DLPFC Visium vs.\ osmFISH & Similar pattern across two spatial-omics technologies.\\
Is the result biologically reasonable? & Predicted domains vs.\ marker genes & Recovery of known layer/domain marker genes.\\
\bottomrule
\end{tabular}
\caption{Planned evaluation. The goal is not only to get a high score, but to
test whether the score depends on real spatial tissue structure.}
\label{tab:evaluation}
\end{table}

% ===========================================================================
\section{Rigor and Biological Interpretation}

\subsection{Testing whether the spatial structure is doing the work}
The most important control will be a graph-removal experiment. I will train the
same GNN under three graph settings: the real spatial $k$NN graph, a shuffled
graph with similar degree but incorrect neighbors, and an empty graph with no
neighbor messages. If the hypothesis is correct, the real graph should perform
best, while the shuffled and empty graphs should move closer to the MLP baseline.
If that does not happen, then the model's advantage is probably not truly
spatial, and I will report that directly.

\subsection{From accuracy to biology}
For each predicted domain I will extract the genes driving predictions and check
whether known cortical-layer marker genes are recovered in DLPFC. I will also
look for spots where the model confidently disagrees with the manual label. Those
spots could be model errors, annotation errors, or small local tissue structures
that deserve closer review. This is important because the project should produce
more than an accuracy number; it should produce biological hypotheses that a
researcher can inspect.

% ===========================================================================
\section{Clinical Translation: Toward Precision Pathology}

The cortical-layer benchmark serves as the proof of concept. The same capability,
accurate and automated domain mapping that stays spatially coherent, is one
requirement for translating spatial transcriptomics to the clinic, particularly
in oncology \cite{cilento2024}.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.62\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Application} & \color{white}\textbf{How domain mapping helps}\\
\midrule
\rowcolor{srirow} Tumor microenvironment & Separate tumor core, invasive front, stroma, and immune niches to study heterogeneity and metastasis.\\
Immunotherapy response & Localize where immune cells sit relative to malignant cells, a spatial correlate of response.\\
\rowcolor{srirow} Neuro-/neurodegenerative disease & Pinpoint which cortical layers carry early molecular pathology (Alzheimer's, schizophrenia).\\
Pathologist support & Annotate complex molecular slides faster and more reproducibly than manual labelling.\\
\bottomrule
\end{tabular}
\caption{The brain benchmark tests the method; precision oncology and pathology
are where it pays off. All predictions are research-only and hypothesis-generating.}
\label{tab:clinical}
\end{table}

% ===========================================================================
\section{Project Strengths and Feasibility}

A few things make the project feasible and worthwhile
(Table~\ref{tab:formula}): it uses real, public data; it has a clear medical
motivation; it compares against tuned non-graph baselines under leakage-safe
splits; it includes a causal check on where the gain comes from; and it ties the
accuracy back to known biology.

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\linewidth} >{\raggedright\arraybackslash}p{0.62\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Criterion} & \color{white}\textbf{Commitment}\\
\midrule
\rowcolor{srirow} Real, sizable public data & LIBD DLPFC: 12 sections, 47k annotated spots, plus a second technology (osmFISH).\\
Clear medical motivation & Spatially-resolved localization of disease pathology; precision oncology bridge.\\
\rowcolor{srirow} Rigorous evaluation & Tuned XGBoost and MLP baselines; leakage-safe splits; multi-seed macro-F1 and ARI.\\
Causal / falsification test & Graph-removal ablation will test whether the GNN depends on real tissue neighborhoods.\\
\rowcolor{srirow} Interpretability $\rightarrow$ discovery & Recover marker genes; surface candidate annotation errors.\\
Transparent reporting & Any weak or negative result will be reported as-is.\\
\bottomrule
\end{tabular}
\caption{Strengths and feasibility of the project.}
\label{tab:formula}
\end{table}

% ===========================================================================
\section{Eight-Week Timeline}

The eight-week plan is designed to start with a reproducible baseline and then
add the graph model, controls, biological interpretation, and a small tumor-ST
extension if time allows (Table~\ref{tab:timeline}).

\begin{table}[H]
\centering\small
\begin{tabular}{>{\raggedright\arraybackslash}p{0.14\linewidth} >{\raggedright\arraybackslash}p{0.78\linewidth}}
\toprule
\rowcolor{sriblue}\color{white}\textbf{Weeks} & \color{white}\textbf{Milestones}\\
\midrule
\rowcolor{srirow} 1--2 & Build the data pipeline for DLPFC and osmFISH; construct spatial graphs; run a small synthetic sanity check.\\
3--4 & Implement expression-only baselines (XGBoost, MLP) and the first GraphSAGE model with shared preprocessing.\\
\rowcolor{srirow} 5 & Run leakage-safe evaluation splits and report macro-F1, ARI, and accuracy across seeds.\\
6 & Run graph controls: true graph, shuffled graph, empty graph, and neighborhood-size sensitivity.\\
\rowcolor{srirow} 7 & Biological interpretation: marker-gene recovery and confident disagreement with manual labels.\\
8 & Prepare figures, tables, manuscript draft, and a reproducible code/data manifest; add a small tumor-ST pilot if time permits.\\
\bottomrule
\end{tabular}
\caption{Schedule for the eight-week SRI cohort. The tumor-ST extension is a
stretch goal after the main brain and osmFISH benchmarks are complete.}
\label{tab:timeline}
\end{table}

% ===========================================================================
\section{Reproducibility and Ethics}

Implementation will use open tools (PyTorch Geometric for the GNN; scanpy and anndata
for spatial-omics I/O; XGBoost). I will save code, configs, random seeds, and a
data manifest, and every reported number will be multi-seed with standard
deviations. No human subjects are involved; all inputs are public, de-identified
post-mortem or model-organism tissue. Predictions are research-only and
hypothesis-generating and do not constitute clinical guidance.

% ===========================================================================
\begin{thebibliography}{9}
\small
\bibitem{cilento2024} Cilento MA, Sweeney CJ, Butler LM. \textit{Spatial
transcriptomics in cancer research and potential clinical impact: a narrative
review}. Journal of Cancer Research and Clinical Oncology, 150(6):296, 2024.
doi:10.1007/s00432-024-05816-0. PMID:~38850363.
\bibitem{maynard2021} Maynard KR, Collado-Torres L, et al. \textit{Transcriptome-
scale spatial gene expression in the human dorsolateral prefrontal cortex}.
Nature Neuroscience, 24:425--436, 2021.
\bibitem{pardo2022} Pardo B, et al. \textit{spatialLIBD: an R/Bioconductor package
to visualize spatially-resolved transcriptomics data}. BMC Genomics, 23:434, 2022.
\bibitem{codeluppi2018} Codeluppi S, et al. \textit{Spatial organization of the
somatosensory cortex revealed by osmFISH}. Nature Methods, 15:932--935, 2018.
\bibitem{long2023} Long Y, et al. \textit{Spatially informed clustering,
integration, and deconvolution of spatial transcriptomics with GraphST}. Nature
Communications, 14:1155, 2023.
\bibitem{dong2022} Dong K, Zhang S. \textit{Deciphering spatial domains from
spatially resolved transcriptomics with an adaptive graph attention auto-encoder
(STAGATE)}. Nature Communications, 13:1739, 2022.
\bibitem{hu2021} Hu J, et al. \textit{SpaGCN: integrating gene expression, spatial
location and histology to identify spatial domains using graph convolutional
networks}. Nature Methods, 18:1342--1351, 2021.
\bibitem{hamilton2017} Hamilton WL, Ying R, Leskovec J. \textit{Inductive
representation learning on large graphs (GraphSAGE)}. NeurIPS, 2017.
\bibitem{velickovic2018} Veli\v{c}kovi\'c P, et al. \textit{Graph Attention
Networks}. ICLR, 2018.
\end{thebibliography}

\end{document}
